---
title: "Transplant Gene Expression Association with Clinical Outcomes"
author: "Patrick Short"
date: "June 1st - June 12th, 2015"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.align = "center", echo=FALSE, warning=FALSE, message=FALSE)
library(png)
library(grid)
library(knitr)
library(ggplot2)
```

```{r scripts and dependencies, include=FALSE}
source("../scripts/eset_tools.R")
source("../scripts/cluster_tools.R")
library(stringr)
library(WGCNA)
```

# Introduction

This R markdown focuses on start to finish analysis of a discovery set of gene expression profiles with associated clinical outcomes. Gene expression sets will be decomposed into gene-modules (collections of genes) using the Weight Gene Coexpression Network Analysis (WGCNA) package. Any correlation between gene module and clinical outcome can be validated statistcally by running the same workflow in an independent set and biologically with GO term enrichment analysis and further experimentation.

Several inputs are necessary for this analysis:
1. CEL file for each patient in the cohort set. These will be checked to contain a Tritan ID (between 600 and 699).
2. Clinical outcomes, or phenotype data in CSV format. One column of this data_set will be matched against the Tritan ID inferred from the CEL file names.
3. Soft thresholding parameter (which should be determined using WGCNA tools).
4. Gene and probeset annotations for Affy Gene ST 1.0 array.

First, we use Robust Multi-array Average (RMA) from the oligo package to build normalized expression sets from the CEL files. We use just batches 2 and 3 (which together contain 32 samples) as QC on revealed large batch effects between batch 1 and batches 2 and 3 (See figure below for PCA plot). We will also filter the genes and probes such that we have a single probe per gene and only use genes with entrez annotations. This will make downstream analysis much more effective - correlation of clinical outcome with an unannotated gene is significantly less useful than an annotated gene.

```{r QC pca plot, echo = FALSE}
img <- readPNG("../results/pca_batch123.png")
grid.raster(img)
```

```{r construct normalized ExpressionSets, include=FALSE, cache=TRUE}

batch2 = list.files(path = "../data/TransplantCELs/batch2_2014", pattern = ".*CEL", all.files = FALSE, full.names = TRUE, recursive = TRUE, ignore.case = TRUE)
batch3 = list.files(path = "../data/TransplantCELs/batch3_2014", pattern = ".*CEL", all.files = FALSE, full.names = TRUE, recursive = TRUE, ignore.case = TRUE)

eset = rma_eset(CEL_list = c(batch2, batch3))

```

```{r filter for genes/probes with Entrez IDs, include=FALSE, cache=TRUE}
# download this file from http://www.affymetrix.com/support/technical/annotationfilesmain.affx
affy_annotation = "../data/HuGene-1_0-st-v1.na35.hg19.transcript.csv"

eset <- gene_filter_eset(eset, affy_annotation)

# extract the expression data and rotate so we have samples (rows) by genes (columns)
expr_data <- t(eset@assayData$exprs)

# change sample names to Tritan IDs
rownames(expr_data) <- str_extract(rownames(expr_data), "[6][0-9][0-9]") # matches 601 - 699

colnames(eset) <- rownames(expr_data)

n_genes = ncol(expr_data)
n_samples = nrow(expr_data)

```

Based on previous analysis, we set the soft thresholding parameter $k = 10$. For the batch 2 and 3 data, this parameter results in a graph topology which is approximately scale free (connections follow a power law), a standard assumption for biological data. We first build a dendrogram (hierarchichal cluster) relating the `r n_genes` in our analysis by topological overlap. The WGCNA dynamic cut algorithm will allow us to partition this dendrogram to define gene modules.

```{r build tree and dynamic cut, echo=FALSE, fig.align='center', fig.width=10, cache=TRUE}
soft_thresh_power = 10

# note - this takes a long time and has been pre-calculated and saved
#clust_out <- build_tree(expr_data, soft_thresh_power)
#gene_tree = clust_out[[1]]
#dissTOM = clust_out[[2]]

# module identification using dynamic tree cut
#modules = cutreeDynamic(dendro = gene_tree, distM = dissTOM,
#deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 30)
#module_colors = labels2colors(modules)

#save(gene_tree, module_colors, file = "../data/geneTree.RData")
load("../data/geneTree.RData")

n_modules = length(unique(module_colors))

```

We can merge modules that are highly correlated - these figures are included at the end of the documents (the ones presented below will be on un-merged modules).

```{r merge highly correlated modules, include = FALSE}

# merge highly correlated modules
MEList = moduleEigengenes(expr_data, colors = module_colors)
MEs = MEList$eigengenes

# cut the tree at 0.20 (80% correlation)
MEDissThres = 0.20

merge = mergeCloseModules(expr_data, module_colors, cutHeight = MEDissThres, verbose = 3)
merged_colors = merge$colors
merged_MEs = merge$newMEs

```

```{r plot merged and unmerged cuts, echo=FALSE, fig.width=8, fig.height = 6}

plotDendroAndColors(gene_tree, cbind(module_colors, merged_colors),
c("Dynamic Tree Cut", "Merged dynamic"), dendroLabels = FALSE, hang = 0.03, addGuide = TRUE, guideHang = 0.05)

```

## Correlating Module Eigengenes with Clinical Phenotypes

Each module (collection of genes) will be represented by a single number which is the first principal component of the module `eigengene' or the linear combination of genes in the module that best captures the intra-module variance. Each sample will have a different value for the eigengene expression which can be correlated with clinical outcomes.

First, we will analyze eigengene correlation against a variety of different clinical outcomes. Many outcomes will be continuous or integral, such as immunophenotyping data on T-cells or cold ischemia time and amenable to pearson correlation (assuming a linear relationship). For categorical or binary data we can use spearman rank correlation. 

We will need to introduce an ordering to some categorical data that is not binary. One such example of an "ordered categorical" outcome will be Cytomegalovirus status in donor and recipient. We will code a donor/recipient negative/negative transplant as 0 (best prognosis), followed by 1 for negative/positive, 2 for positive/positive, and 3 for positive/negative (worst prognosis). We clearly do not expect this relationship to be linear, simply monotonically increasing, so a spearman rank correlation is the most appropriate choice. 

We focus on the following clinical outcomes:

```{r clinical outcomes, echo=FALSE}

phenotypes <- c("RRT_length", "cold_ischemia_time", "DGF", "immunological_risk", "sensitized", "rejection", "rejection_1wk", "rejection_2wk", "rejection_1month", "rejection_3month", "rejection_1yr", "viral_complication", "bacterial_complication", "fungal_complication", "number_HLAmm", "number_DRmm", "CMV_status", "CMV_IgG", "K_preTx", "IgG_level", "GFR_preTx", "CRP", "t_cell_immunophenotyping<sup>+</sup>")

data_type <- c("numeric", "numeric", "binary", "binary", "binary", "binary","numeric", "numeric", "numeric", "numeric", "numeric", "binary", "binary", "binary", "numeric", "numeric", "ordered_categorical", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric")

correlation <- c("pearson", "pearson", "spearman", "spearman", "spearman", "spearman", "spearman", "spearman", "spearman", "spearman", "spearman","spearman", "spearman", "spearman", "pearson" , "pearson", "spearman", "pearson", "pearson", "pearson", "pearson", "pearson", "pearson")

tab = cbind(phenotypes, data_type, correlation)[order(correlation), ]

knit_print(kable(tab))

```

<sup>+</sup>T cell immunophenotyping is not a single measure, but a large set of quantitative measures describing the T cell population by flow using the same aliquot of blood used to generate the gene expression data used in this analysis.

It is important to note that a pearson correlation is not used for all of the numeric phenotypes. A pearson correlation carries the assumption of a linear relationship and therefore may not be appropriate for an outcome such as the number of rejection episodes within the first month ($rejection_1month$) where we would not expect changes in gene expression to correlate linearly with the number of episodes, only monotonically (the spearman assumption).

Using a false discovery rate (FDR) cut-off of 0.05, we can visualize all of the continuous phenotypes for which there is at least one significantly correlated gene module. Correlation between gene module and phenotype ranges from blue (perfectly anti-correlated) to white (uncorrelated) to red (perfectly correlated).

The phenotypes labels for both continous and discrete measures are clustered by similarity across samples.

```{r remove phenotypes with missing data and cluster, include=FALSE}

continuous_pheno <- read.csv("../data/transplant_continuous_phenotypes.csv")
discrete_pheno <- read.csv("../data/transplant_discrete_phenotypes.csv")

# sort on patient number then remove
continuous_pheno <- continuous_pheno[match(colnames(eset), continuous_pheno$patient_number), -1]
discrete_pheno <- discrete_pheno[match(colnames(eset), discrete_pheno$patient_number), -1]

# we will remove any phenotypes that have >75% missing data
continuous_pheno <- continuous_pheno[ , colSums(is.na(continuous_pheno)) < n_samples*0.75]
discrete_pheno <- discrete_pheno[ ,colSums(is.na(discrete_pheno)) < n_samples*0.75]

# reorder discrete and continuous pheno by correlating columns
cont_tree = hclust(as.dist(cor(data.matrix(continuous_pheno), use = "p")), method="average")
disc_tree = hclust(as.dist(cor(data.matrix(discrete_pheno), use = "p", method = "spearman")), method = "average")

continuous_pheno <- continuous_pheno[ ,cont_tree$order]
discrete_pheno <- discrete_pheno[ ,disc_tree$order]

```


```{r correlate eigengenes with continuous phenotypes, include=FALSE}

MEs0 = moduleEigengenes(expr_data, module_colors)$eigengenes
MEs = orderMEs(MEs0)

fdr = 0.05

# Display the correlation values within a heatmap plot
pheno_heat_fig(expr_data, module_colors, continuous_pheno, 
               fname = "../results/continuous_corr_heatmap.pdf", 
               signif_only = TRUE, fdr = fdr, colors_only = TRUE)

# note that labeledHeatmap does not render within knitr - need to save and load as a chunk

```

<center><img src="../results/continuous_corr_heatmap.pdf"></center>

We can do the same for discrete clinical outcomes. In this case, we plot all of the phenotypes and mark those with significant correlations (at false discovery rate of 0.01) with an asterisk.

```{r discrete phenotypes, include = FALSE}
##### discrete traits #####

pheno_heat_fig(expr_data, module_colors, discrete_pheno, 
               fname = "../results/discrete_corr_heatmap.pdf", 
               signif_only = FALSE, fdr = fdr, colors_only = TRUE)


save(eset, module_colors, MEs, continuous_pheno, discrete_pheno, file = "../results/eset_batch23.RData")

```

<center><img src="../results/discrete_corr_heatmap.pdf"></center>

## Gene Ontology Term Enrichment

If these gene module to phenotype associations are real, then we expect the gene modules to be biologically informative in some way. One way to test that gene modules represent biologically meaningful groups of genes is to test for enrichment of Gene Ontology terms within each module.


```{r GO enrichment by module, include=FALSE, cache=TRUE}

all_entrez_ids = eset@featureData@data$EntrezID
GOenr = GOenrichmentAnalysis(module_colors, all_entrez_ids, organism = "human", nBestP = 10);

# get the table for best 10 terms over all three GO categories
# table contains enrichment p value ($BonferoniP), ontology term ($termOntology), etc.
GO_table = GOenr$bestPTerms[["BP, CC, MF"]]$enrichment

GO_by_module = split(GO_table, f = GO_table$module)
n_enriched_GO = sapply(GO_by_module, function(m) sum(m$BonferoniP < 0.05))

```

Out of the `r length(n_enriched_GO)` gene modules, `r sum(n_enriched_GO > 0)` have at least one GO term which is significantly enriched against the background (Bonferroni adjusted p-value < 0.05). Taking the ten terms per module with the lowest p-value, we can see that many of the modules have several enriched terms. Most of these terms are similar and imply a shared function or pathway of the gene module. A few of the modules have no apparent enrichment of biological terms. In fact, we expect this to be the case for the grey module (to which WGCNA assigns all genes left out of clusters).

```{r plot number of enriched GO terms, echo=FALSE, fig.width = 8, fig.height = 6}

mod_colors = as.character(names(n_enriched_GO))
counts = as.integer(n_enriched_GO)

qplot(x = mod_colors, y = counts, geom = "bar", stat="identity", fill = mod_colors) +
  scale_fill_manual(values = mod_colors) + 
  labs(x = "Module Colors", 
       y = "Number of Enriched GO Terms (out of top 10)",
       title = "Most Gene Modules are Highly Enriched for GO terms.") +
  theme(axis.text.x = element_text(colour="black",size=16,angle=90,hjust=.5,vjust=.5),
        plot.title = element_text(colour="black",size=20,angle=0,hjust=.5,vjust=.5),
        axis.title =element_text(colour="black",size=16,angle=0,hjust=.5,vjust=.5)) + 
  theme(panel.grid.minor = element_blank()) +
  guides(fill = FALSE)


```

There are `r sum(n_enriched_GO == 0)` gene modules with no highly enriched GO terms (`r mod_colors[counts == 0]`). It is reassuring that we see the grey module in this list as we expect this module to consist of genes that did cluster into any module.

Merged module phenotype correlation heat maps:

```{r merged module heat maps, include=FALSE}
pheno_heat_fig(expr_data, merged_colors, continuous_pheno, 
               fname = "../results/continuous_corr_merged_heatmap.pdf", 
               signif_only = TRUE, fdr = fdr)

pheno_heat_fig(expr_data, merged_colors, discrete_pheno, 
               fname = "../results/discrete_corr_merged_heatmap.pdf", 
               signif_only = FALSE, fdr = fdr)

```

<center><img src="../results/continuous_corr_merged_heatmap.pdf"></center>
<center><img src="../results/discrete_corr_merged_heatmap.pdf"></center>


